---
title: "Differential_diff_in_diff"
author: "Joey Herrera"
date: "4/18/2021"
output: pdf_document
---
# Differential timing difference in difference R script

Needs:

- ai cross sectional units - 16 metro regions in Texas
- at time periods - 14 from 2005 to 2019
- BDD*Dit beta coefficients for the 2x2 estimate and the associated treatment dummy

The first thing to do is establish the year when each metropolitan region in Texas recieved their cultural district. This will allow me to control for fixed effects via differential timing and establish a never treated group for regions who are still without a cultural district.

Years when cultural districts were established: 
Source of reference: https://www.arts.texas.gov/initiatives/cultural-districts/map-of-cultural-districts/?doing_wp_cron=1618786448.0652329921722412109375#:~:text=Dallas%20%E2%80%93%20Deep%20Ellum%20Cultural%20District


Amarillo - 2005 - http://artsinamarillo.com/ -1
Austin - 2013 - https://www.sixsquare.org/about/ - 2
Beaumont - 2019 - https://downtownbeaumont.org/welcome-to-dbcad/ - 1
Brownsville - NA
Dallas/ Fort Worth - 1984 - https://www.dallasartsdistrict.org/about/ - 4
El Paso - 2011 - https://mcad.elpasotexas.gov/arts-district - 1
Houston - 1999 - https://www.eastenddistrict.com/about/ - 7
Laredo - NA
Lubbock - circa 2010 - https://www.facebook.com/lbkculturaldistrict/ (contact) - 1
Midland - NA
Odessa - NA
San Angelo - 2019 - http://www.sanangeloculturaldistrict.org/faqs/index (contact) - 1
San Antonio - 2011 - https://kingwilliamculturalartsdistrict.com/about - 3
Tyler - NA

When creating the dummy variabled for the year the cultural districts were established, I will use zero before 2005 and 9999 as the never-treated regions.

```{r setup, include=FALSE}
# code a dummy variable for the treatment group
TX_housing = TX_housing %>%
  mutate(treat = ifelse(YEAR >= 2013,1,0))

TX_housing = TX_housing %>%
  filter(RegionName != "San Angelo")

```

## R Markdown



```{r }
logit_housing <- glm(treat ~ mean_age +  prop_male + prop_hisp  + 
                   mean_INCTOT + mean_FTOTINC , family = binomial(link = "logit"), 
                 data = TX_housing)
logit_housing

```



```{r, echo=FALSE}

# Fit propoensity score to quad model
prop_housing <- TX_housing %>% 
  mutate(pscore = logit_housing$fitted.values)

# mean pscore control
pscore_control <- prop_housing %>% 
  filter(treat == 0) %>% 
  pull(pscore) %>% 
  mean()

# mean pscore treated
pscore_treated <- prop_housing %>% 
  filter(treat == 1) %>% 
  pull(pscore) %>% 
  mean()
```

```{r}
# Create histograms
# logit quad control

prop_housing %>% 
  filter(treat == 0) %>% 
  ggplot() +
  geom_histogram(aes(x = pscore))

# logit quad treatment
prop_housing %>% 
  filter(treat == 1) %>% 
  ggplot() +
  geom_histogram(aes(x = pscore))
```

```{r}
# trimming propensity score for logit and OLS data
# logit model (quadratic)
prop_housing_trim <- prop_housing %>% 
  filter(!(pscore >= 0.9)) %>% 
  filter(!(pscore <= 0.1))

# mean pscore control
pscore_control__trim <- prop_housing_trim %>% 
  filter(treat == 0) %>% 
  pull(pscore) %>% 
  mean()
# 0.3661185

# mean pscore treated
pscore_treated_trim <- prop_housing_trim %>% 
  filter(treat == 1) %>% 
  pull(pscore) %>% 
  mean()
#0.6254476
```

```{r}

### First difference for the logit cubic model
prop_housing_trim_treat = prop_housing_trim
prop_housing_trim_control = prop_housing_trim

prop_housing_trim_treat = prop_housing_trim_treat %>%
  filter(!is.na(avg_price))
prop_housing_trim_control = prop_housing_trim_control %>%
  filter(!is.na(avg_price))

# Filter avg_price to only include people in the treatment group
prop_housing_trim_treat %>% 
  filter(treat == 1) %>% 
  summarise(avg_price)

# Filter and average on the treatment effect
mean1 <- prop_housing_trim_treat %>% 
  filter(treat == 1) %>% 
  pull(avg_price) %>% 
  mean()

# save the 
prop_housing_trim_treat$y1 <- mean1

# Average treatment effect on the untreated
prop_housing_trim_control %>% 
  filter(treat == 0) %>% 
  summarise(avg_price)

mean0 <- prop_housing_trim_control %>% 
  filter(treat == 0) %>% 
  pull(avg_price) %>% 
  mean()

prop_housing_trim_control$y0 <- mean0

ate_logit <- unique(prop_housing_trim_treat$y1 - prop_housing_trim_control$y0)
# 18315.67
prop_housing_trim <- prop_housing_trim %>% 
  filter(treat == 1) %>% 
  select(-y1, -y0)
```

```{r}
##### Question 3 weighted difference in difference regression
### Cubic logit model
N <- nrow(prop_housing_trim)
# Manual with non-normalized weights using trimmed data for the cubic logit model
prop_housing_trim <- prop_housing_trim %>% 
  mutate(d1 = treat/pscore,
         d0 = (1-treat)/(1-pscore))

s1 <- sum(prop_housing_trim$d1)
s0 <- sum(prop_housing_trim$d0)

prop_housing_trim <- prop_housing_trim %>% 
  mutate(y1 = treat * avg_price/pscore,
         y0 = (1-treat) * avg_price/(1-pscore),
         ht = y1 - y0)

prop_housing_trim %>% 
  pull(ht) %>% 
  mean()
```

